Relationships in the extremes and their influence on competition and coexistence

### 2022-07-13 paper/Paper.Rnw
editts from this morniing - addressing comments from both LMH and SPE about clarity of the same paragraph in theory section

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. We develop each step of our new theory for a general 
two-species model (below), and also specifically for the lottery model of the Introduction 
(SI section \ref{SI-SIsect:TheoryForLottery}).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: the storage effect. 
So we begin by defining the storage effect. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption %can be circumvented
is not neccesary, %we make the assumption because 
it simplifies the presentation of new ideas.

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between invader and resident species mean growth rates, when $i$ is the invader.
Specifically, storage effect for species $i$ is 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
%Though some readers may find it slightly confusing to simultaneously use the variable $E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like all operators, $\E(\cdot)$ is rendered upright.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, 
(see eqn. (13) in Ellner (2019),
whereas ATAs are %defined in a covariation context
strictly about covariation. 
Therefore, the storage effect is the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is analogous to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- the storage effect. 
Only storage effect involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is the storage effect. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction. One typically %(but see below) 
must examine this quantity for each species, $i$,
to obtain complete information about coexistence. To pursue goal G2, below, we will quantify
the magnitude of the new ATA term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms.

>>>>

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. We develop each step of our new theory for a general 
two-species model (below), and also specifically for the lottery model of the Introduction 
(SI section \ref{SI-SIsect:TheoryForLottery}).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show
that, for purely temporal variation, only one mechanism relates to tail associations: the storage effect. 
So we begin by defining the storage effect. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption %can be circumvented
is not neccesary, %we make the assumption because 
it simplifies the presentation of new ideas.

In MCT, coexistence is assessed by the ability to increase when rare, i.e., 
the mean ``growth when rare'' [GWR] of species $i$, 
quantified by the weighted difference between mean growth rates when species $i$, the invader, is at negligible abundance 
and species $j$, the resident (the other species) is at stochastic steady state. The GWR is computed for each species assessed as the invader of a resident community, and only when the difference is positive for all is coexistence possible. 
The weighted difference, $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, can be decomposed into coexistence mechanisms. 
A subscript $i \bs i$ indicates a quantity computed for species $i$ when that species is at negligible abundance, 
and a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent. 
The \emph{scaling factor} $q_{ij}$ weights the relative sensitivities to competition of the two species, 
see \cite{Chesson_1994}, \cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two alternative definitions of $q_{ij}$ that we use.

The storage effect %for species $i$ 
is the contribution of covariation between $E$ and 
$C$ to a difference between invader and resident species mean growth rates  \cite{Ellner_2016}. %, when $i$ is the invader.
Specifically, storage effect for species $i$ is 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
%Though some readers may find it slightly confusing to simultaneously use the variable $E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like all operators, $\E(\cdot)$ is rendered upright.

@@

The other mechanisms in MCT that apply in our modelling context of purely temporal variation
make no reference to covariation between quantities, 
(see eqn. (13) in Ellner (2019),
whereas ATAs are %defined in a covariation context
strictly about covariation. 
Therefore, the storage effect is the only mechanism of coexistence
for which ATAs are relevant. Mathematical details supporting this claim are in SI section 
\ref{SI-SIsect:other_mechanisms}, but we here provide a conceptual understanding. \cite{Ellner_2016} and
\cite{Ellner_2019} provide a decomposition of the coexistence metric $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$,
which equals GWR of species $i$. The decomposition, which we use below, is analogous to and motivated by the 
original decomposition of \cite{Chesson_1994}; the decomposition is how the mechanisms of 
coexistence operating in a given case are illuminated. The decomposition is 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{(EC)},\label{eq:previous_decomp}
\end{equation}
where: $\Delta_i^0$ is a baseline;
$\Delta_i^E$ is a contribution to coexistence of environmental variation;
$\Delta_i^C$ is a contribution to coexistence of variation in competition;
$\Delta_i^{(E\shp C)}$ is a contribution to coexistence of having simultaneous variation in both environment and competition, but not including the effects 
of covariation in these quantities; and
$\Delta_i^{(EC)}$ is a contribution to coexistence of $EC$ covariation itself -- the storage effect. 
Only storage effect involve $EC$ covariation,
so the only mechanism of coexistence influenced by ATA between $E$ and $C$ is the storage effect. Combining our decomposition 
(\ref{eq:storage_effects_further}) with the decomposition (\ref{eq:previous_decomp}) of \cite{Ellner_2019} gives 
\begin{equation}
\olr_{i \bs i} -q_{ij}\olr_{j \bs i}=\Delta_i^0+\Delta_i^E+\Delta_i^C+\Delta_i^{(E\shp C)}+\Delta_i^{[E \nat C]} + \Delta_i^{[EC]},\label{eq:full_decomp}
\end{equation}
which is the extension of MCT which was goal G1 of the Introduction. 
%One typically %(but see below) must examine this quantity for each species, $i$,to obtain complete information about coexistence. 
To pursue goal G2, below, we will quantify
the magnitude of the new ATA term, $\Delta_i^{[EC]}$, relative to the magnitudes of the other terms.
